---
title: "Cheatsheet for Programming in R"
output: html_notebook
date: '2022-04-01'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### R vectors

A vector is a data structure that can hold multiple objects of the same data type. For example, `c(45, 25, 89, 10)` creates a numerical vector. To access elements, use the `[ ]` brackets with indexing starting from 1.

```{r}
my_string_vector <- c("this", "is", "an", "example", "vector")
my_string_vector[3] # returns "an"
 
my_boolean_vector <- c(TRUE, FALSE, FALSE)
my_numerical_vector <- c(0.4, 0.9, 1, 0.45, 1.2, 0.33)
 
# R will force the same type even if you input different types to c()
my_vector <- c("word", 45, 12, FALSE)
# result: a vector of strings "word" "45" "12" "FALSE"
```

### R Lists

In R, a list can store a variety of data types and data structures in a single variable. This means a list can hold individual strings and numbers, but also vectors and lists.

To access a single element, use `[[ ]]` double brackets. A list is useful for organization because you can name the elements inside it and select them using $

```{r}
my_list <- list("Elephant", FALSE, 900, 80.3, list("pencil", "pens"))
my_list[[4]] # returns 80.3
 
my_longer_list <- list(misc = my_list, notes = c("g", "b", "d", "g"))
 
my_longer_list[[1]]
# returns the contents of my_list
my_longer_list$misc
# returns the contents of my_list
my_longer_list$notes
# returns "g" "b" "d" "g"
```
### R Matrices
To create one, use the `matrix()` function with a vector representing the data (R will interpret whether it is repeating or non-repeating), nrow equal to the number of rows, and ncol equal to the number of columns.

To access a single element inside a matrix called ex_matrix, you can look up the item in the rth row and the cth column like so: `ex_matrix[r, ]`.

To access the entire rth row, you can refer to it like so: `ex_matrix[r, c]`. To access the cth column, use `ex_matrix[ , c]`.

```{r}
# will create a 3-by-3 matrix with the vectors being filled column-wise (default)
my_matrix <- matrix(c("a", "b", "e", "k", "e", "w", "g", "x", "t"), nrow = 3, ncol = 3)
# result:
# "a" "k" "g"
# "b" "e" "x"
# "e" "w" "t"
single_element <- my_matrix[1, 3]
# returns "g"
 
second_row <- my_matrix[2,]
# returns: "b" "e" "x"
second_col <- my_matrix[,2]
# returns: "k" "e" "w"

```

### R Control Flow
Control flow involves the program deciding which code to execute. The decision-making is established through conditional statements, i.e. if, else if, and else. Each condition should compute to a logical TRUE or FALSE. You can use comparison operators like !, & and | to combine logical values.

```{r}
if (condition_to_check) {
   # execute code and don't check any more conditions
} else if (other_condition_to_check & and_this_condition_to_check) {
   # execute code only if both are true and don't check any more conditions
} else if (either_this_condition | or_this_condition ) {
   # execute code if either condition is true and don't go to else
} else {
   # the default code if none of the conditions above are true
}
```

### Loops in R
```{r}
# how to define a for loop
for (loop_variable in sequence) {
    # code to repeat
}
 
# how to define a while loop
while (condition_to_check_every_iteration) {
    # code to repeat
}
```

### Functions in R
To apply a function onto all the elements in a data structure, we can use one of the apply functions with that function’s name.

 - `apply()` is used on a two-dimensional data structure, and margin value indicates either row or column.
 - `sapply()` is used on any data structure and returns a vector or matrix depending on the number of dimensions.
 - `lapply()` is used on any data structure and always returns a list.

```{r, eval=FALSE}
# the function constructor
my_function <- function(){}

# define a function
function_name <- function(parameter_1, parameter_2, ....) {
   # do something with parameter_1 and parameter_2
   return(some_value)
}

apply(data_structure, margin_value, function_name)
sapply(data structure, function_name)
lapply(data_structure, function_name)
```

#### Sample with replacement
```{r}
sample(x = 1:6, size = 2, replace = TRUE)
```
### Saving Data
```{r}
# writing file to csv
write.csv(file_name, file = "new_file_name.csv", row.names = FALSE)

```


